Manuscript_draft

### 2021-01-06 manuscript_draft.Rmd
Add text from Mike's emails

@@

```{r define captions to be used in methods results and discussion, echo=FALSE}

# Captions (put these in order)

  # Introductory 
  appNums(name="paiTable",caption="Number of growing seasons between each measurement period",display=FALSE) # Table showing number of growing seasons between each measurement period
  appNums(name="standVol.plot",caption="Plot total volume for five measurement periods",display=FALSE) # Plot volume figure
  appNums(name="standBA.plot",caption="Plot basal area for five measurement periods",display=FALSE)  # Plot basal area figure
  appNums(name="standQMD.plot",caption="Plot quadratic mean diameter for five measurement periods",display=FALSE) # QMD figure  
  appNums(name="standK.plot",caption="Plot tree density for five measurement periods", display=FALSE) # SPH figure

  # Diameter class analysis
  figNums("diamVol.plot", caption="Mean plot volume distribution by diameter class and treatment, for two periods.  Black line shows total volume.  Volume calculated on trees meeting 17.5cm DBH limit.  Spruce and fir contribution to volume by diameter class are shown with bar plots.  Douglas-fir volume contribution by diameter class is not shown.",display="FALSE")

  figNums("diamBA.plot", caption="Mean plot basal area distribution by diameter class and treatment, for two periods.  Black line shows total basal area.  Spruce and fir contribution by diameter class are shown ith bar plots.  Douglas-fir contribution by diameter class is not shown.",display="FALSE")

  figNums("diamK.plot", caption="Mean tree density distribution by diameter class and treatment, for two periods.  Black line shows total tree density.  Spruce and fir contribution by diameter class are shown ith bar plots.  Douglas-fir contribution by diameter class is not shown.",display="FALSE")

@@

>>>>

# To-do list

Comments from Mike - we can put text below each comment to describe how it was addressed, and then delete comments for any issues once they are resolved.

From November 6 e-mail:

Use of terminology

  *   The term "units" is used in several places, where it seems that you mean "treatments" (or treatment types, treatment unit).

  *   The use of the term "plot" or "plots" is used appropriately as far as I can see.

  *   The term "fir" is used to describe subalpine fir or 'balsam' in several parts of the text, including possibly figure and table captions. I know what species you are referring to, of course, but others, especially cursory readers, may misconstrue it. The colloquial use of 'fir" is spread across both Abies and Douglas-fir. "Balsam" as a term has well-documented ambiguities, and while Canadians tend to use balsam for Abies, Americans often call Abies "firs", as in the various species of true fir. I would recommend using Abies as the shorthand descriptor in the text and captions for subalpine fir.  It is a short word, and we have only 1 Abies species at this site. Also, my experience in the on-line-search literature world and with things like Google Scholar is that Abies is a much better and more diagnostic search word for this subject matter than fir or balsam. "Spruce-Abies" is very unambiguous as a descriptor for this forest type.

  *   Rate of change (increase, decrease) in vol, BA, and sph: In both in your stats tables and figure captions, and in the text narrative, ensure that what you are describing is clearly and accurately described as either:  (a) the incremental change in values between two measurement periods, or (b)  the difference in absolute value in certain parameters (e.g. standing volume) between two measurement periods, or between two treatments. Also, the annualized growth rate of something averaged over a measurement period (e.g. mean PBAI, like BA growth in m2/yr over a period) is different than the rate of change in a growth rate over time (e.g. is a growth rate increasing or decreasing).  I got confused a few times on this issue.

Commentary on sections and sub-sections

Section 3.1.1 BA increment from 1992 to 2019

  *   Great: Very clear and straightforward. i.e. - log-transformed treatment (initial BA) was a very statistically significant predictor of tree response across dbh classes for both Sx and Bl.
  *   Don't know what all the random effects outputs are. I am sure they are meaningful, but I just dont know what they mean.
  *   Number of observations of 510: Should ensure it is clear that only trees > 7.5 cm dbh are counted.
  *   Would be good to define the dbh class limits somewhere. In Appendix 6 table, we describe the class mid-point.
  *   For Appendix 6, I assume that the null hypothesis for the ratio is = 1.

Section 3.1.2 Mortality

  1.  I am thinking that mortality should / could be analysed as a frequency response - i.e. - expected vs actual distribution of mortality by size class and/or species. e.g. a Chi-squared or allied type of analysis. Null hypothesis is that mortality of stems is randomly distributed across different sizes and species of trees. H1 is that it is non-random.
  2.  I'd suggest that we don't compare total mortality (e.g # of stems) across treatment units, but rather frequency relative to initial number of stems (i.e. - at beginning of observation period). That way we normalize mortality rates relative to initial stem densities. Otherwise, results could be mis-leading - e.g. Control treatments could have more mortality in overall sph, simply because they have more initial stems.
  3.  ACTION ITEM FOR ME:  I need to check my original datasets and data sorting iterations to see how I handled mortality observations for the 1992 and 1994 plot remeasurements.
  4.  Note for you:  All of the mortality you've analysed and identified so far is mortality that is incremental following 92/94, so it is still solid analysis data, just with that qualification. I will clarify the 92/94 mortalilty observations for the Methods section; we did observe dead trees at these times, but these mortality pre-dated our treatments and/or first measurements. Anyway, probably not a big deal, but this needs examination and clarification to explain this. If necessary, I will quantify observations of dead trees observed on these years, because they are actually > 0.

Testing for density-dependence of mortality responses... some ideas or hypotheses..  Note that f (___) means "a function of"...

  *   MORT RATE of a size class = f (Sum of BA in that plot > than that size class).
Would assume asymmetric competition where only bigger trees exert competition on smaller trees.
  *   MORT RATE of a size class = f ( Sum of BA in that plot that is either > or < that size class).
Assumes symmetric competition where both big and small trees exert competition on their neighbors.

Caveats:   (a) our plots and tree distributions might be too spatially variable to get a reliable test of these hypotheses, and (b) BA would be the BA at the beginning of a measurement period. And (c) a tree that died of suppression would probably show reduced growth rates prior to death; but trees that had abiotic damage and mortality - e.g. - snow or wind breakage - might not. Complicated stuff !!

3.2.1  Estimating tree height from diameter

  *   Really solid stats and rationale for a single ht:dbh model. Great !
  *   What is really interesting and useful about this is that the ht:dbh relationship is seemingly statistically independent of treatment or, that is, initial density,  A very useful simplifying assumption proven for future modellers of these stands. We should highlight this in either our results or discussion.

3.2.2 Plot means

  *   See notes above about use of term "units" vs treatment (or similar).
  *   Clarify effects of treatment and/or species on growth rate vs standing vol / BA at each measurement period.
  *   A Figure 7b or equivalent could visually present changes in growth rate over time. (e.g. previous graphs we've developed)
  *   Under Observations, # 2 - the 2nd sentence needs work to edit and clarify.

3.2.3 Stand attributes by diameter class

  *   We should define what we mean by the term inverse J or inverse J-shaped dbh distribution. Just for clarity.
  *   Should we quantitatively describe these distributions? Don't have to, just asking the question.
  *   Is the merch limit for min merch tree size 18.5 cm dbh, or 17.5 cm dbh? I remember the latter from various cruises and cruise compilations.
  *   There are two Figure 2's, one is for vol and one is for sph.
  *   Clarify in figures and text that sph is based on trees 7.5 cm dbh and greater. Increases in sph can be from ingress of trees growing from size classes below this limit.
  *   A suggestion:  Have a table comparing total volume to merch volume by treatment, etc?  Would be useful for some readers, and communicate MAI and volume growth rates too.

3.2.4  Species composition

  *   First sentence:  Clarify what you mean by "organized similarly". Does this refer to frequency distribution ? What constitutes similarity?

# Introduction

# Methods

@@

@@

### Mortality

From Mike's e-mail:

Last winter, as well as keeping reference copies of the initial “raw” dataset following data entry and related stuff, I kept copies each stage of data cleaning and prep (about 10 iterations).

We did have “dead tree” observations in the first ~ 2 year of monitoring. The key thing to see is : if I can determine is when the trees were dead, and/or when death was initiated, and why.  
Timing of tree death and observations of causal factors of tree death in these first two years will help inform our interpretation and data handling of these observations.

If trees were dead or moribund at the first “initial” observation or two (as soon after the logging as possible), this is – arguably -  different than mortality that occurs over time in response to the basal area treatment and/or could be considered to be density- or treatment-dependent in some way.

The 1992 data was in mid-March really early after logging, and before snowmelt (in retrospect not a great idea but we had resources available before the end of the fiscal year at the time). The 1994 data was from when the plots were re-visited under snow-free conditions. We did a thinning and sanitation culling of trees in the treatment units and PSP’s in 1994 to remove logging-damaged trees and most deciduous, to bring the units and PSP’s down to their final treatment BA. 1994 remeasurements were completed after that.

A few plots, as you know, were established in 1994 not 1992.

Interpretive implications:

1.	The very-early observed mortality may be a result of pre-existing stand conditions not the created, treatment-induced stand conditions. “stochastic” vs potentially treatment induced. 
2.	It can be argued that trees dead at time of treatment or very soon  thereafter were never taking up growing space in the post-harvest stand, and therefore would not contribute to the initial “green” basal area going forward. 
3.	We do have 1992 data in most plots, but it was 1994 when the BA in each trmt unit and embedded PSP was finalized. 1994 is the most homogeneous start point for all PSP’s.

Regarding tree mortality in the period from 1992 to 1994 at Summit Lake EP 1162:

I reviewed our data regarding this, any relevant project records, and my own recollection of our monitoring of tree mortality during this period. Here is my train of thought, and what I have concluded on this topic:

1992 Observations and PSP establishment limitations

Trees that were dead at the time of the 1992 PSP establishment were not tagged or recorded. The rationale for this was that these trees were dead and therefore not taking up up growing space in the stands. Dead trees in 1992 would have been either snags / dead trees pre-existing the harvest, or trees fatally damaged or killed by logging activities (e.g. – trees with stems snapped and crowns broken off, no live limbs) or trees smashed to the ground by felled trees).

Summer 1994 sanitation treatment and Fall 1994 PSP Remeasurement

EP 1162 treatment units, including the PSP area contained therein, were assessed in Spring 1994 to review stand density / basal area, and stand condition, in order to:  
a)	identify and mark-to-cut residual trees that were considered too badly scarred or damaged to constitute good growing stock for trial purposes, 
b)	where needed, to lower the initial BA closer to the target level, 
c)	to either fell or girdle birch trees, especially in the PSP’s themselves, but usually also within the TU as a whole, and 
d)	to fell any dead, dangerous, and/or moribund trees within the treatment units and PSP’s. 
Generally most if not all of these trees felled in the sanitation felling were understory or subdominant trees, and occasionally a lower codominant tree. Tagged trees felled by the sanitation felling were documented on field sheets and our Excel datasets, so we can retrieve the species and size of these trees as needed. However, sanitation-felled trees included both live, moribund, and dead trees – generally all poorer quality or vigor though.

After felling, the felled trees were bucked to a 2 to 3 m length so as to lay these felled trees to as close to the ground as possibly, maximize decay, and minimize physical obstructions within the TU’s and PSP’s.

As a result of the sanitation felling, potentially some trees which died or were damaged between 1992 and Summer 1994 were felled without recording these mortality events and precise tree vigor beforehand. Some minor mortality occurred between 1992 and 94, but was obscured by the post-harvest sanitation treatment.

The 1994 Remeasurement occurred in Fall 1994, with a few plot remeasurements occurring in Winter 1994/95 (i.e. no intervening growing season).

For scientific rigour, we should not state that there was zero mortality between 1992 and 1994. Rather, I suggest, it is more accurate to state that data on tree mortalilty between 1992 and 1994 was not collected, and that the Summer 1994 sanitation felling treatment did not allow accurate documentation of tree mortality in this initial period on a retrospective basis.

We don’t, therefore, have data on mortality between 1992 and 1994, though my recollection is that it was minor in nature.

The difference in basal area between 1992 and 1994 is predominantly a product of the sanitation felling of some basal area, with minor natural mortality.

Mortality occurring between Fall 1994 and Fall 1997 remeasurements was captured accurately by these two sequential remeasurements.

“The difference in basal area between 1992 and 1994 is a result of basal area reductions through the sanitation felling of some stems and some minor natural mortality, combined with basal area increases through the growth of remaining live stems.”

A review of the various plots indicates that between 1992 and 1994, some PSP’s decreased in BA (as sanitation removals and minor mortality exceeded growth of live stems), while others increased in BA as tree growth exceeded BA removals.
